Reasoning Graphically about Quantum Computation

### 2008-03-02 papers/reasoning-graphically/paper.tex
working...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1392 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[runningheads]{llncs}
%\usepackage[dvips]{color}
%\usepackage{multicol}
%\usepackage{tabls}
%\usepackage{ttbox}
\usepackage{allmtt}
%\usepackage{pstricks,pst-node,pst-tree} % PS Tricks for diagrams
%\usepackage{proof2}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{stmaryrd}
%\usepackage{mbboard}
\usepackage{xspace}
\usepackage{diagrams}
\usepackage{proof2}

\usepackage{url}
\urldef{\mailsa}\path|l.dixon@ed.ac.uk|
\urldef{\mailsb}\path|ross.duncan@comlab.ox.ac.uk|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

\begin{abstract}
  Systems of quantum computation are typically represented by large
  matrix compositions. However, recent graph-based formalisms offer an
  approach which exposes the structure of these systems in a clearer
  and more accessible way. Although these provide a significantly
  simpler way of reading and presenting quantum computations, manual
  manipulation of such graphs is slow and error prone. We present a
  formalism that supports a mechanised reasoning about such
  graphs. This involves a compositional account of graph rewriting
  that preserves the underlying categorical semantics. We present a
  system, which has been implemented, with a fixed logical kernel and
  support for derived rules, which allows the model of quantum
  computation to be specified declarativley.  Soundness of the system
  is proved with respect to the underlying model and completeness is
  an open problem.

@@

History of diagrammatic representation going back to ???

CAn formalise quantum mechanics in an abstract setting using compact
or $\dag$-compact categories.

@@

We then introduce a concrete set of generators and equations for use
in quantum computation.  We need rewrites to capture the non-logical
axioms of this structure.   It turns out that we can most easily express
these equations using an infinite family of ``spider rules'';  this
motivates the  more general notion of pattern graphs.

next section introduces the  notion of pattern graph, which is
essentially a graph structured

@@

A \emph{directed graph}\footnote{
Equivalently:  a directed graph is a functor $G$ from  $\bullet
\pile{\rTo\\\rTo} \bullet$ to \catSet;  a graph morphism is then a natural
transformation  $f: G \Rightarrow H$.
} consists of a 4-tuple $(V,E,s,t)$ where $V$
and $E$ are sets, respectively of \emph{vertices}\footnote{We will use
  the words ``vertex'' and ``node'' interchangeably.} and \emph{edges},
and $s$ and $t$ are maps 
\begin{diagram}
  E & \pile{\rTo^{\qquad s\qquad}\\\rTo_t} & V
\end{diagram}
which we call \emph{source} and \emph{target}.  Define sets
$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ of \emph{incoming}
and \emph{outgoing} edges at a vertex $v$.  
The \emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
\sizeof{\text{out}(v)}$.

Given graphs $G$ and $H$, a \emph{graph morphism} $f : G\to H$ consists of
functions $f_E : E_G \to E_H$ and $f_V:V_G\to V_H$ such that:
\begin{gather}
  s_H\circ f_E = f_V \circ s_G,\label{eq:graph-hom1}\\
  t_H\circ f_E = f_V \circ t_G\label{eq:graph-hom2}.
\end{gather}

Let $f: G \to H$ be  graph morphism and  let $V' \subseteq V_G$. We say
that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
that $f_E(e') = e$.  We call $f$ an \emph{open embedding} for $V'$ if:
\begin{enumerate}
\item $f_E$ is injective;
\item $f_V$ restricted to $V'$ is injective; and,
\item $f$ is strict for $V'$.
\end{enumerate}
The intuition behind this definition is that the subgraph of $G$
determined by $V'$ should be preserved exactly by $f$, whereas the
other vertices may be identified.

Given two graphs $G,H$, a pair of partial maps $r_V: V_G
\rightharpoonup V_H$, $r_E: E_G \rightharpoonup E_H$ is called a
partial graph morphism if whenever $r(e)$ is defined then $r(s_G(e))$
and $r(t_G(e))$ are defined, and the restriction of $r_E$ to its
preimage satifies equations \eqref{eq:graph-hom1} and
\eqref{eq:graph-hom2} above.

@@

\begin{definition}
A strict symmetric monoidal  category
\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact 
closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
object $A^*$, and morphisms
\begin{gather*}
  d_A : I \to A^* \otimes A\\
  e_A : A \otimes A^* \to A
\end{gather*}
such that
\begin{align}
  A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A
  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A & = \id{A} \label{eq:comcl1}\\
  A^* \iso I \otimes A^* \rTo^{ d_A \otimes \id{A^*}} A^* \otimes A
  \otimes A^* \rTo^{\id{A^*} \otimes e_A} A^* \otimes I \iso A^* & =
  \id{A^*} \label{eq:comcl2}
\end{align}
\end{definition}

@@

\begin{figure}
  \figureline

  The $(\cdot)^\dag$ functor gives a vertical symmetry to the
  category, hence for every equation we have a second equation
  obtained by flipping the diagram upside down.  In additon, we have a
  ``colour duality'': each equation shown here gives rise to
  second, which is obtained by exchanging the two colours.  (In fact
  this duality is derivable from the equations for involving $H$.)
  \begin{description}
  \item[Comonoid Laws] 
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
      \inlinegraphic{3em}{comonoid-assoc2}
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-unit1}  &=\;&
      \inlinegraphic{3em}{comonoid-unit2}  &=\;&
      \inlinegraphic{3em}{comonoid-unit3} 
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-comm1}  &=\;&
      \inlinegraphic{3em}{comonoid-comm2}
    \end{array}
    \]
  \item[Isometry, Frobenius, and  Compact Structure]
    \[
    \begin{array}{cccccccccccccccc}
      \inlinegraphic{3.5em}{isometry1}  &=\;&
      \inlinegraphic{3.5em}{isometry2}  
      &\qquad\qquad &
      \inlinegraphic{3em}{frobenius1}  &=\;&
      \inlinegraphic{3em}{frobenius2}  
      &\qquad\qquad&
      \inlinegraphic{2.5em}{compact1}  &=\;&
      \inlinegraphic{1.3em}{compact2}  
    \end{array}
    \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
    \end{array}
    \]
  \item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
    \inlinegraphic{1.5em}{bialgebra2}$;  then:
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} 
  \end{array}
  \]
\item[Group Actions]
  \[
  \begin{array}{cccccccccccccccccccccc}
    \inlinegraphic{2.5em}{group-int1} &=\;&
    \inlinegraphic{2.5em}{group-int2} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int3} &=\;&
    \inlinegraphic{2em}{group-int4} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int5} &=\;&
    \inlinegraphic{1.5em}{group-int6} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int7} &=\;&
    \inlinegraphic{1.5em}{group-int8} 
  \end{array}
  \]
    \item[$H$ Properties]
    \[
    \begin{array}{cccccccccccccccccccc}
      \left(\inlinegraphic{1.5em}{H}\right)^\dag &=\;&
      \inlinegraphic{1.5em}{H} 
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq1} &=\;&
      \inlinegraphic{2.5em}{Heq2}
    \end{array}
    \]
  \item[Colour Duality]
        \[
    \begin{array}{cccccccccccccccccccc}
      \inlinegraphic{2.5em}{Heq-delta} &=\;&
      \inlinegraphic{1.5em}{reddelta}
      &\qquad\qquad&
      \inlinegraphic{2em}{Heq-epsilon} &=\;&
      \inlinegraphic{1.5em}{redepsilon}
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq-alpha} &=\;&
      \inlinegraphic{1.5em}{redalpha}
    \end{array}
    \]

@@

  \caption{Graphical Equations for Quantum Systems}
  \label{fig:graph-quant-eqns}
  \figureend
\end{figure}

@@

\noindent However, composition with semicircle line(s) on the domain
or co-domain allows a proof involving one of the above to easily
derive a corresponding one for the other two. Proving any of the
following allows a trivial derivation of the others:

@@

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we need a
representation that abstracts over the boundary nodes membership in
the domain or co-domain. In this section we formalise such a notion
and refer to the generalised boundary nodes as \emph{variable
  nodes}. This gives rise to a \emph{variable-node graphs}, which are
graphs with variable nodes instead of boundary edges.

Below, we give a more formal and constuctive account the semnatics by
introducing a notion of an \emph{open embedding} and note that this
corresponds to the intuitive idea of graph-matching. We then go on and
introduce a way to join graphs with variable-nodes and note it's
relationship to composition on the underlying concrete graphs.

\subsection{Semantics}

\subsection{Matching as an Open Embedding}

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea with

\begin{definition}
  Matching is charcterised by an \emph{open embedding} which is strict
  for the non-variable nodes.
\end{definition}

The intuition of variable nodes is that they can replaced by any be
nodes of the right type that are themseleves within any graph.  A
graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs, denoted by
$\vinterp{G}$.

% Thus, unlike nodes in conrete a graph, variable nodes do not define an
% ordering on the edges that are incident. Instead, the edges that touch
% a variable node are simply a set. The type of a variable nodes is thus
% defined as a multiset rather than a tensor. Any tensor type can be
% lifted to a multiset type in the obvious manner, which we denote by
% the operation {\tt mset\_of\_tensor}. For example, {\tt
%   mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
% say that a tensor-type $T$ is an instance of a multiset type $M$,
% written $T \in M$, when $M \subseteq
% \mathtt{mset\_of\_tensor}(T)$. The intuition here is that when a
% variable node is instantiated to some other node, it's type will have
% at least the currently attached edges.

@@

\begin{figure}[t]
%  \scalebox{1.0}{\includegraphics{images/node-var-instance.eps}}
*** Add figure *** 
\label{node-variable-instances-fig}\caption{The concrete graphs $a$,
  $b$, and $c$ are instances of $x$, a graph with variable nodes.}
\end{figure}

We refer to members of this set as instances.

For example, in Figure~\ref{node-variable-instances-fig}, the graphs
$a$, $b$ and $c$ are instances of the graph $x$. A graph with variable
nodes, and without boundary nodes, is called a \emph{variable-node
  graph}.

\subsubsection{Completness and Adequacy.}

Variable-node graphs are complete in the sense that for every concrete
graph there is a variable node graph that can be instantiated to it:
$\forall g.\,\exists G.\,g \in \vinterp{G}$. The witness for this
proof is made by simply replacing all bounary nodes with variable
nodes.

@@

In this section we outline how the formalism we have presented for
graph patterns can be used to provide a \emph{meta-level} framework
for reasoning about models of compact closed categories. Following the
terminology of logical frameworks such as Isabelle~\cite{isabelle}, we
call the specification provided by the underlying model an
\emph{object-level} graph formalism.

An object-level formalisation defines a set of rules which are treated
as the axioms for the system. The object-level implementation also
defines the data at the nodes and edges as well as corresponding
data-matching behaviour. The meta-level provides machinery to
manipulate graphs and derive new equations. In this section we
describe the meta-level framework and prove its adequacy for
rewriting.

@@

Axioms defined by an object-level model, as well as derived rules in
our framework, are declarative in the sense that they are pairs of
pattern graph. This pair represents the left and right hand sides of
an equation.

Substituon with a rule then involves replacing a subgraph that matches
the left hand side with the right hand side. To do this there has to
be a isomorphism between variable nodes in the left and right hand
subgraphs. More concretely, there must be a unique naming of the
variable nodes in the left hand side such that each name occurs in a
unique variable node on the right hand side. Given a matching on the
left hand side, the target subgraph can be replaced with right hand
side connected by the corresponding variable nodes. This provides
rewriting for pattern graphs.

\subsubsection{Adequacy}

A an equation that involves a graph pattern on one side

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[runningheads]{llncs}
%\usepackage[dvips]{color}
%\usepackage{multicol}
%\usepackage{tabls}
%\usepackage{ttbox}
\usepackage{allmtt}
%\usepackage{pstricks,pst-node,pst-tree} % PS Tricks for diagrams
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{stmaryrd}
%\usepackage{mbboard}
\usepackage{xspace}
\usepackage{diagrams}
\usepackage{proof2}
\usepackage{url}

\urldef{\mailsa}\path|l.dixon@ed.ac.uk|
\urldef{\mailsb}\path|ross.duncan@comlab.ox.ac.uk|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

\begin{abstract}
  Recent graph-based formalisms of quantum computation provide an
  abstract and symbolic way to represent and simulate the
  computations. However, manual manipulation of such graphs is slow
  and error prone. We present a formalism that supports mechanised
  reasoning about such graphs. This involves a compositional account
  of graph rewriting that preserves the underlying categorical
  semantics. We present a generic system with a fixed logical kernel
  for representing compact closed categories. This provides a
  declarative account which supports derived rules. We illustrate the
  framework by instantiating it for the graphical language of quantum
  computation.

@@

%History of diagrammatic representation going back to ???

Can formalise quantum mechanics in an abstract setting using compact
or $\dag$-compact categories.

@@

We then introduce a concrete set of generators and equations for use
in quantum computation.  We need rewrites to capture the non-logical
axioms of this structure.   It turns out that we can most easily express
these equations using an infinite family of ``spider rules'';  this
motivates the  more general notion of graph patterns.

%next section introduces the  notion of pattern graph, which is
%essentially a graph structured

@@

A \emph{directed graph}\footnote{ Equivalently: a directed
  graph is a functor $G$ from $\bullet \pile{\rTo\\\rTo} \bullet$ to
  \catSet; a graph morphism is then a natural transformation $f: G
  \Rightarrow H$.  } consists of a 4-tuple $(V,E,s,t)$ where $V$ and
$E$ are sets, respectively of \emph{vertices}\footnote{We will
  use the words ``vertex'' and ``node'' interchangeably.} and
\emph{edges}, and $s$ and $t$ are maps
\begin{diagram}
  E & \pile{\rTo^{\qquad s\qquad}\\\rTo_t} & V
\end{diagram}
which we call \emph{source} and \emph{target}.  The definitions
$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ express the
\emph{incoming} and \emph{outgoing} edges at a vertex $v$.  The
\emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
\sizeof{\text{out}(v)}$.

To distinguish between elements of different graphs, we will use the
subscript notation $G = (V_G,E_G,s_G,t_G)$. Given graphs $G$ and $H$,
a \emph{graph morphism} $f : G\to H$ consists of functions $f_E : E_G
\to E_H$ and $f_V:V_G\to V_H$ such that:

\begin{gather}
  s_H\circ f_E = f_V \circ s_G,\label{eq:graph-hom1}\\
  t_H\circ f_E = f_V \circ t_G\label{eq:graph-hom2}.
\end{gather}

Let $f: G \to H$ be graph morphism and  let $V' \subseteq V_G$. We say
that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
that $f_E(e') = e$.

\begin{definition}
\label{open-embedding-def}
We call $f$ an \emph{open embedding} for $V'$ if:
\begin{enumerate}
\item $f_E$ is injective;
\item $f_V$ restricted to $V'$ is injective; and,
\item $f$ is strict for $V'$.
\end{enumerate}
\end{definition}

The intuition behind this definition is that the subgraph of $G$
determined by $V'$ should be preserved exactly by $f$; it should
contain $G$ without any extra incident egdes into the verticies $V'$
whereas other vertices may be identified and may contain additional
verticies.

Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
\rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
partial graph morphism if whenever $r_E(e)$ is defined then
$r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
$r_E$ to its preimage satifies equations \eqref{eq:graph-hom1} and
\eqref{eq:graph-hom2} above.

@@

\begin{definition}
\label{compactcat-def}
A strict symmetric monoidal  category
\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact 
closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
object $A^*$, and morphisms
\begin{gather*}
  d_A : I \to A^* \otimes A\\
  e_A : A \otimes A^* \to A
\end{gather*}
such that
\begin{align}
  A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A
  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A & = \id{A} \label{eq:comcl1}\\
  A^* \iso I \otimes A^* \rTo^{ d_A \otimes \id{A^*}} A^* \otimes A
  \otimes A^* \rTo^{\id{A^*} \otimes e_A} A^* \otimes I \iso A^* & =
  \id{A^*} \label{eq:comcl2}
\end{align}
\end{definition}

@@

\begin{figure}[p]
  \figureline
  \begin{description}
  \item[Comonoid Laws] 
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
      \inlinegraphic{3em}{comonoid-assoc2}
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-unit1}  &=\;&
      \inlinegraphic{3em}{comonoid-unit2}  &=\;&
      \inlinegraphic{3em}{comonoid-unit3} 
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-comm1}  &=\;&
      \inlinegraphic{3em}{comonoid-comm2}
    \end{array}
    \]
  \item[Isometry, Frobenius, and  Compact Structure]
    \[
    \begin{array}{cccccccccccccccc}
      \inlinegraphic{3.5em}{isometry1}  &=\;&
      \inlinegraphic{3.5em}{isometry2}  
      &\qquad\qquad &
      \inlinegraphic{3em}{frobenius1}  &=\;&
      \inlinegraphic{3em}{frobenius2}  
      &\qquad\qquad&
      \inlinegraphic{2.5em}{compact1}  &=\;&
      \inlinegraphic{1.3em}{compact2}  
    \end{array}
    \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
    \end{array}
    \]
  \item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
    \inlinegraphic{1.5em}{bialgebra2}$;  then:
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} 
  \end{array}
  \]
\item[Group Actions]
  \[
  \begin{array}{cccccccccccccccccccccc}
    \inlinegraphic{2.5em}{group-int1} &=\;&
    \inlinegraphic{2.5em}{group-int2} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int3} &=\;&
    \inlinegraphic{2em}{group-int4} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int5} &=\;&
    \inlinegraphic{1.5em}{group-int6} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int7} &=\;&
    \inlinegraphic{1.5em}{group-int8} 
  \end{array}
  \]
    \item[$H$ Properties]
    \[
    \begin{array}{cccccccccccccccccccc}
      \left(\inlinegraphic{1.5em}{H}\right)^\dag &=\;&
      \inlinegraphic{1.5em}{H} 
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq1} &=\;&
      \inlinegraphic{2.5em}{Heq2}
    \end{array}
    \]
  \item[Colour Duality]
        \[
    \begin{array}{cccccccccccccccccccc}
      \inlinegraphic{2.5em}{Heq-delta} &=\;&
      \inlinegraphic{1.5em}{reddelta}
      &\qquad\qquad&
      \inlinegraphic{2em}{Heq-epsilon} &=\;&
      \inlinegraphic{1.5em}{redepsilon}
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq-alpha} &=\;&
      \inlinegraphic{1.5em}{redalpha}
    \end{array}
    \]

@@

  \caption{Graphical Equations for Quantum Systems. The $(\cdot)^\dag$
    functor gives a vertical symmetry to the category, hence for every
    equation we have a second equation obtained by flipping the
    diagram upside down.  In additon, we have a ``colour duality'':
    each equation shown here gives rise to second, which is obtained
    by exchanging the two colours. The colour duality is derivable
    from the equations for involving $H$.}
  \label{fig:graph-quant-eqns}
  \figureend
\end{figure}

@@

\noindent However, composition with semi-circles ($d$ and $e$ from
Definition~\ref{compactcat-def}) on the co-domain or domain allows an
equation involving one of the above to easily lead to a derivation
corresponding either of the other two. For instance, proving any of
the following allows a trivial derivation of the others:

@@

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we
formalise a representation that abstracts over the boundary nodes
membership in the domain or co-domain. This gives rise to a
\emph{variable-node graphs}, in which boundary nodes have been
generalised to \emph{variable nodes}. The intuition of variable nodes
is that they can replaced by concrete nodes in some graph. We now
define the semantics for variable-node graphs and then go on and
introduce a way to join them.

% relationship to composition on the underlying concrete graphs.

\subsection{Semantics by Matching as an Open Embedding}

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea of graph with variable
nodes occuring within another graph.

\begin{definition}
  \emph{Matching} is an open embedding
  (Definition~\ref{open-embedding-def}) which is strict for the
  non-variable nodes.
\end{definition}

A graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs, denoted by
$\vinterp{G}$. The interpretation is simply the ways a graph with
variable nodes can be matched to a concrete graph. We refer to members
of this set as instances. Matching is also well-defined between
variable-node graphs, in which case we denote that $G$ macthes $H$ by
$G \leq_v H$.

\begin{theorem}
  $G \leq_v H \Leftrightarrow \vinterp{G} \superseteq \vinterp{H}$. 
\end{theorem}

% \begin{figure}[t]
% %  \scalebox{1.0}{\includegraphics{images/node-var-instance.eps}}
% *** Add figure *** 
% \label{node-variable-instances-fig}\caption{The concrete graphs $a$,
%   $b$, and $c$ are instances of $x$, a graph with variable nodes.}
% \end{figure}

% For example, in Figure~\ref{node-variable-instances-fig}, the graphs
% $a$, $b$ and $c$ are instances of the graph $x$. A graph with variable
% nodes, and without boundary nodes, is called a \emph{variable-node
%   graph}.

%% TYPES 
% Thus, unlike nodes in conrete a graph, variable nodes do not define an
% ordering on the edges that are incident. Instead, the edges that touch
% a variable node are simply a set. The type of a variable nodes is thus
% defined as a multiset rather than a tensor. Any tensor type can be
% lifted to a multiset type in the obvious manner, which we denote by
% the operation {\tt mset\_of\_tensor}. For example, {\tt
%   mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
% say that a tensor-type $T$ is an instance of a multiset type $M$,
% written $T \in M$, when $M \subseteq
% \mathtt{mset\_of\_tensor}(T)$. The intuition here is that when a
% variable node is instantiated to some other node, it's type will have
% at least the currently attached edges.

@@

% \subsubsection{Completness and Adequacy.}

% Variable-node graphs are complete in the sense that for every concrete
% graph there is a variable node graph that can be instantiated to it:
% $\forall g.\,\exists G.\,g \in \vinterp{G}$. The witness for this
% proof is made by simply replacing all bounary nodes with variable
% nodes.

@@

In this section we describe how the garph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the specification
provided by the underlying model an \emph{object-level} graph
formalism. In particular, an object-level formalisation defines a set
of rules which are treated as the axioms for the system. It also
defines the data at the nodes and edges as well as corresponding
data-matching behaviour. For it's part, the meta-level provides
machinery to manipulate graphs and derive new rules. In this section
we describe the meta-level framework and prove its adequacy for
rewriting.

@@

Axioms defined by an object-level model, as well as derived rules in
our framework, are pairs of pattern graph. The pair represents the
left and right hand sides of an equation. Rules are declarative in
that they denote set of concrete equational rules.

Substituon with a rule involves replacing a subgraph that matches the
left hand side with the right hand side. To do this there has to be a
isomorphism between variable nodes in the left and right hand
subgraphs. More concretely, there must be a unique naming of the
variable nodes in the left hand side such that each name occurs in a
unique variable node on the right hand side. Given a matching on the
left hand side, the target subgraph is replaced with the right hand
side while keeping the same instantiations for the isomorphic variable
nodes.

Rules also contain a partial 1-1 mapping between !-boxes on the left
and right hand sides. The intuition for this mapping is that the
unfolding used when matching, for a !-box on the left, is applied to
the mapped !-box on the right before replacement.

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example,
Figure~\ref{spider-law2-fig} illustrates the Spider Law as a rule in
the general framework. The partial 1-1 mapping for !-boxes is
represented by !-boxes having the same name on the left and right of a
rule. Similarly, the isomorphism between variable nodes is captured by
the set of variable node names on the left and right hand side being
equal. There is an additional restriction that comes from the
interplay between !-boxes and variable nodes. When a variable node
appears within a !-box on one side of a rule it must appear under
mapped !-box on the other side. This enforces that rules cannot change
the type of a graph pattern.

This
provides rewriting for pattern graphs.

A an equation on a graph pattern is
